---
title: "Analyzing Eviction Lab Data, 2000-2018"
author: "Angel Mendiola Ross"
date: '2023-06-17'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_knit$set(root.dir = '/Users/angelmr/solano_sr/evictions/')
```

## Set up

```{r}
### Clear global environment
rm(list=ls())

library(pacman)
pacman::p_unload(all)

pacman::p_load(
  tidyverse, #dplyr, readr, etc.
  data.table, #fread() 
  foreign, #load data types including stata .dta files 
  magrittr, #%<>% operator
  skimr, #for summerising
  tidyverse,
  readxl #to read excel files
)
```

## Evictions Data Analysis

This analysis is based on data from 2000-2018 provided by the Eviction Lab <https://data-downloads.evictionlab.org/#data-for-analysis/>. It draws on two of their data sets: "county_court-issued_2000_2018.csv" (which provides court-issued data for counties) and "tract_proprietary_valid_2000_2018.csv" (which provides proprietary data for census tracts). I reduced both raw csvs just to counties in California to reduce file size. All other manipulation and analysis done within this file. I did not use the "county_proprietary_valid_2000_2018.csv" file because there was no data for Solano County.

Data for Solano County across both data sets is only available from 2010 to 2017, so I restrict analysis to these years.

I will also bring in census data to illustrate change in tenure and renter/homeowner affordability measures from 2000 to 2017.

```{r}
# load data
county <- read_csv("data/raw/county_court-issued_2000_2018.csv")
tract <- read_csv("data/raw/tract_proprietary_valid_2000_2018.csv")

# reducing datasets to 2010-2017 because this is what is available for Solano
county %<>%
  subset(year>2009)
tract %<>%
  subset(year>2009)
```

First I will compute rates per 100 renting households for the measures in the county data set (similar to how they are reported in the tract-level data set).

```{r}
county %<>%
  mutate(filing_rate = (filings_observed/renting_hh)*100,
         threat_rate = (hh_threat_observed/renting_hh)*100)
```

Now I will get a few demographic and economic variables for the county from the ACS.

```{r}
# getting demographic data back to 2010
library(tidycensus)
census_api_key("1ce78664b294208f88f53c49746f378a51893b62", overwrite = TRUE)

# acs data - getting data through 2021 in case I find more recent eviction data
years <- lst(2010, 2011, 2012, 2013, 2014, 2015, 2016, 2017, 2018, 2019, 2020, 2021)

dp10 <- load_variables(2010, "acs5", cache = TRUE)

race_acs <- c(
  total_pop = "B03002_001",
  total_hhs = "B25003_001",
  owner_hhs = "B25003_002",
  renter_hhs = "B25003_003",
  black_hhs = "B25003B_001",
  black_owner = "B25003B_002",
  black_renter = "B25003B_003",
  asian_hhs = "B25003D_001",
  asian_owner = "B25003D_002",
  asian_renter = "B25003D_003",
  latino_hhs = "B25003I_001",
  latino_owner = "B25003I_002",
  latino_renter = "B25003I_003",
  nhw_hhs = "B25003H_001",
  nhw_owner = "B25003H_002",
  nhw_renter = "B25003H_003",
  u_own = "B25106_002",
  own_lt20k = "B25106_003",
  own_rb_lt20k = "B25106_006",
  own_2035k = "B25106_007",
  own_rb_2035k = "B25106_010",
  own_3550k = "B25106_011",
  own_rb_3550k = "B25106_014",
  own_5075k = "B25106_015",
  own_rb_5075k = "B25106_018",
  own_75kpls = "B25106_019",
  own_rb_75kpls = "B25106_022",
  own_0k = "B25106_023",
  u_rent = "B25106_024",
  rent_lt20k = "B25106_025",
  rent_rb_lt20k = "B25106_028",
  rent_2035k = "B25106_029",
  rent_rb_2035k = "B25106_032",
  rent_3550k = "B25106_033",
  rent_rb_3550k = "B25106_036",
  rent_5075k = "B25106_037",
  rent_rb_5075k = "B25106_040",
  rent_75kpls = "B25106_041",
  rent_rb_75kpls = "B25106_044",
  rent_0k = "B25106_045",
  rent_norent = "B25106_046",
  med_hh_inc = "B25119_001",
  med_hh_inc_own = "B25119_002",
  med_hh_inc_rent = "B25119_003"
)

acs_county <- map_dfr(
  years,
  ~ get_acs(
    geography = "county",
    variables = race_acs,
    state = "CA",
    year = .x,
    survey = "acs5"
  ),
  .id = "year"
)

# deleting moe from acs dataframe
acs_county %<>%
  select(-c(moe))

# transforming
acs_wide_cty <- acs_county %>%
                    distinct() %>%
                    spread(variable, estimate)
```

Now finalizing the data set to join with evictions data.

```{r}
# computing the share of renter (and homeowner) households who are cost burdened
acs_wide_cty %<>%
  mutate(cb_owner = own_0k + own_rb_lt20k + own_rb_2035k + own_rb_3550k + own_rb_5075k + own_rb_75kpls,
         p_cb_owner = cb_owner/(u_own),
         cb_renter = rent_0k + rent_rb_lt20k + rent_rb_2035k + rent_rb_3550k + rent_rb_5075k + rent_rb_75kpls,
         p_cb_renter = cb_renter/(u_rent-rent_norent),
         prent = renter_hhs/total_hhs,
         pown = owner_hhs/total_hhs,
         prent_blk = black_renter/black_hhs,
         prent_his = latino_renter/latino_hhs,
         prent_nhw = nhw_renter/nhw_hhs,
         prent_asian = asian_renter/asian_hhs)

# reducing cols for simplicity
acs_wide_cty <- acs_wide_cty[c(1:18,30,43:57)]

# saving the census data
save(acs_wide_cty, file = "data/acs_county_data.RData")

# merging with county evictions data
county %<>% rename(GEOID = fips_county)
county$GEOID <- as.character(county$GEOID)
county$year <- as.character(county$year)
county %<>%
  mutate(GEOID = paste0("0", GEOID))

county %<>%
  select(-state)
county %<>%
  select(-fips_state)

df <- left_join(x = county,
                y = acs_wide_cty,
                by = c("year", "GEOID"))

df %>%
  arrange(desc(filing_rate))

df %>%
  subset(county %in% c("Solano County", "Alameda County", "San Mateo County", "Contra Costa County")) %>%
  ggplot(aes(x = year, y = filing_rate)) +
  geom_bar(stat="identity") +
  ggtitle("Eviction Filing Rate by County, 2010-2017") +
  xlab("Year") + ylab("Eviction Filing Rate per 100 Renting Households") +
  facet_wrap(~ county) +
  theme_classic()

df %>%
  subset(county %in% c("Solano County")) %>%
  ggplot(aes(x = year, y = filing_rate)) +
  geom_bar(stat="identity") +
  ggtitle("Eviction Filing Rate in Solano County, 2010-2017") +
  xlab("Year") + ylab("Eviction Filing Rate per 100 Renting Households") +
  theme_classic()

df %>%
  subset(county %in% c("Solano County", "Alameda County", "San Mateo County", "Contra Costa County")) %>%
  select(county,year,filing_rate,p_cb_renter,cb_renter,renter_hhs,prent)

df %>%
  subset(county %in% c("Solano County")) %>%
  ggplot(aes(x = year, y = p_cb_renter)) +
  geom_bar(stat="identity") +
  ggtitle("Rent Burdened Households in Solano County, 2010-2017") +
  xlab("Year") + ylab("% of Renters Spending More Than 30% of Income on Rent") +
  facet_wrap(~ county) +
  theme_classic()
```

Now preparing the census tract data file.

```{r}
acs_tract <- map_dfr(
  years,
  ~ get_acs(
    geography = "tract",
    variables = race_acs,
    state = "CA",
    year = .x,
    survey = "acs5"
  ),
  .id = "year"
)

# deleting moe from acs dataframe
acs_tract %<>%
  select(-c(moe))

# transforming
acs_wide_trct <- acs_tract %>%
                    distinct() %>%
                    spread(variable, estimate)

# saving the census data
save(acs_wide_trct, file = "data/acs_tract_data.RData")
```

Now getting the measures of interest and joining with tract-level eviction dataset.

```{r}
# computing the share of renter (and homeowner) households who are cost burdened
acs_wide_trct %<>%
  mutate(cb_owner = own_0k + own_rb_lt20k + own_rb_2035k + own_rb_3550k + own_rb_5075k + own_rb_75kpls,
         p_cb_owner = cb_owner/(u_own),
         cb_renter = rent_0k + rent_rb_lt20k + rent_rb_2035k + rent_rb_3550k + rent_rb_5075k + rent_rb_75kpls,
         p_cb_renter = cb_renter/(u_rent-rent_norent),
         prent = renter_hhs/total_hhs,
         pown = owner_hhs/total_hhs,
         prent_blk = black_renter/black_hhs,
         prent_his = latino_renter/latino_hhs,
         prent_nhw = nhw_renter/nhw_hhs,
         prent_asian = asian_renter/asian_hhs)

# reducing cols for simplicity
acs_wide_trct <- acs_wide_trct[c(1:18,30,43:57)]

# saving the census data
save(acs_wide_trct, file = "data/acs_tract_data.RData")

# merging with county evictions data
tract %<>% rename(GEOID = fips)
tract$GEOID <- as.character(tract$GEOID)
tract$year <- as.character(tract$year)
tract %<>%
  mutate(GEOID = paste0("0", GEOID))

tract %<>%
  select(-state)

df_tract <- left_join(x = tract,
                y = acs_wide_trct,
                by = c("year", "GEOID"))

df_tract %>%
  arrange(desc(filing_rate))

df_tract %>%
  subset(county=="Solano County") %>%
  arrange(desc(filing_rate)) %>%
  select(tract, county, year, filings, filing_rate, p_cb_renter, prent_blk, prent_his)
```

Saving out final county-level and tract-level data files.

```{r}
# data was showing up weird in Tablaeu so getting rid of NANs
df_tract %<>% mutate_all(~ifelse(is.nan(.), NA, .))

# county first
save(df, file = "data/df_county.RData")
write_csv(df, file = "data/df_county.csv", na = "")

# now tract
save(df_tract, file = "data/df_tract.RData")
write_csv(df_tract, file = "data/df_tract.csv", na = "")
```
